Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-16 ms.tex
Adding radial profile of kph for the variable source.  Adding descriptions of Test 12, the X-ray test, and the radiation pressure test.

Figure \ref{fig:test7_1} shows cuts of density, temperature, neutral
fraction, and pressure in a line connecting the source and the clump
center at $t = 1$, 10, and 50 Myr.  At 1 Myr, the ionization front has
propagated through the left-most 500 pc of the clump.  This heated gas
is now overpressurized, as seen in the pressure plot in Figure
\ref{fig:test7_1}, and then expands into the ambient medium.  This
expansion creates a photo-evaporative flow, seen in many star forming
regions \citep[e.g. the Carina Nebula][]{Ref?} as stars irradiate
nearby cold, dense overdensities.  These flows become evident in the
density at 10 Myr, seen both in the line cuts and slices (Figure
\ref{fig:test7_2}).  These outflows have temperatures up to 50,000 K.
At this time, the front has progressed about halfway through the
clump, if one inspects the neutral fraction.  However the high energy
photons have heated all but the rear surface of the clump.  At the end
of the test ($t = 50$ Myr), only the core and its associated shadow is
neutral, as seen in Figure \ref{fig:test7_3}.  The core has been
compressed by the surrounding warm medium, thus causing the higher
densities seen at $t = 50$ Myr.  The non-spherical artifacts on the
inner boundary of the warm outermost shell are caused by the initial
discretization of the sphere, as discussed in \S\ref{sec:test3}.

@@

Some astrophysical systems produce collimated radiation either
intrinisically by relavistic beaming or by an optically-thick torus
absorbing radiation in the equatorial plane.  The latter case would be
applicable in a subgrid model of an AGN or a protostar, for example.
Simulating collimated radiation with ray tracing is trivially
accomplished by only initializing rays that are within some opening
angle $\theta_c$.

@@

We illustrate the expansion of the \ion{H}{2} region created by the
beamed radiation in Figure \ref{fig:test11}.  Before 3 Myr, the
\ion{H}{2} region is conical and contained within the dense clump,
depicted in the $t = 0.1$ Myr snapshot of the system.  At this time,
the ionization front is transitioning from R-type to D-type in the
transverse direction of the cone.  This can be seen in the minute
overdensities on the \ion{H}{2} transverse surface.  When it breaks
out of the overdensity, it creates a champange flow, where the
ionization front transitions back to a weak R-type front.  The cloud
surface is a constant-pressure contact discontinuity (CD) with a
density jump of 100.  After the front heats the gas at the CD, there
exists a pressure difference of $\sim 100$.  In response, the high
density gas accelerates into the ambient medium and heats it to $3
\times 10^4$ K.  Additionally a rarefaction wave travels towards the
clump center.  At later times, the transverse D-type front continues
through the clump, eventually creating a disk-like structure at the
final time.  The polar champange flows proceed to flow outwards,
creating a density shell with a diffuse ($10^{-28}$ \cubecm) and warm
(5000 K) medium in its wake.

@@

\begin{figure}[t]
  \epsscale{0.6}
  \plotone{fig/test12_kph}
  \caption{\label{fig:test12_kph} Test 12 (Time variations of the
    source luminosity).  Slice of the photo-ionization rate
    \kph~through the origin.  The source has a duty cycle of 0.5 Myr,
    and the box has a light crossing time of 3.3 Myr.  The shells of
    high \kph~originate from radiation that was emitted when the
    source was at its peak luminosity, illustrating the
    time-dependence of the radiative transfer equation.}
  \epsscale{1}
\end{figure}

Our implementation retains the time derivative of the radiative
transfer equation (Eq. \ref{eqn:rteqn}) if we choose a constant ray
tracing timestep, which saves the photon packages between timesteps if
the $c\;dt_{\rm P} < L_{\rm box}$.  This effect only becomes apparent
when the variation timescale of the point source is smaller than the
light crossing time of the simulation.  Furthermore, the timestep
should resolve the variation timescale by a few times.  To test this,
we can use an exponentially varying source with some duty cycle.  In a
functional form, this can be described as
%
\begin{equation}
  \label{eqn:test13}
  L(t) = L_{\rm max} \times \exp[A(t_f/t_0 - 1)]
\end{equation}
%
where $t_f = 2 \times |t - t_0 \times \mathrm{round}(t/t_0)|$, $t_0$
is the duty cycle, and $A=4$ controls the width of the radiation
pulse.  To illustrate the variations in the photo-ionization rates,
shown in Figure \ref{fig:test12_kph}, due to a varying source, we
remove any dependence on the medium by considering an optically-thin
uniform density $\rho 10^{-4}~\cubecm$.  We take $L_{\rm box} = 1$
Mpc, which has a light crossing time of 3.3 Myr.  A source is placed
at the origin with $L_{\rm max} = 10^{55}$ ph s$^{-1}$ and $t_0 = 0.5$
Myr.  We use a radiative transfer timestep of 50 kyr to resolve the
duty cycle by 10 timesteps.

@@

As the \ion{H}{2} region grows the magnetized medium, shown in Figures
\ref{fig:test14_1} and \ref{fig:test14_2}, it transforms from
spherical to oblate as it is magnetically confined in directions
perperdicular to the magnetic field.  This occurs at $t > 0.5$ Myr
because the magnetic pressure exceeds the thermal pressure, and the
gas can only flow along field lines.  \citeauthor{Krumholz07_ART}
observed some carbuncle artifacts along the ionization front; whereas
we see smooth density gradients, which is most likely caused by our
more diffusive choice of the HLL Riemann solver when compared to Roe's
Riemann solver used in \citet{Krumholz07_ART}, which also uses PLM as
a reconstruction method.  The evolution of the magnetic field lines
evolve in a similar manner as their results.

@@

\begin{figure}[t]
  \epsscale{0.6}
  \plotone{fig/xray_profiles.eps}
  \caption{\label{fig:xray_fig} Radial profiles of temperature and
    ionized fraction showing the effects of secondary ionizations from
    a mono-chromatic 1 keV spectrum.  The high energy photons can
    ionize multiple hydrogen atoms, increasing the ionized fraction.
    In part, less radiation goes into thermal energy, lowering the
    temperature.}
  \epsscale{1}
\end{figure}

@@

@@

\begin{figure}[t]
  \plottwo{fig/noRP_profiles.eps}{fig/RP_profiles.eps}
  \caption{\label{fig:rp_profile150} (a) No radiation pressure.
    Radial profiles of (clockwise from top left) density, temperature,
    radial velocity, and neutral fraction.  Time units in the legned
    are in kyr.  (b) Radial profiles with radiation pressure.  The
    momentum transferred to the gas drives out the gas at higher
    velocities than without radiation pressure.  Afterwards the
    central region is underpressurized, and the gas infalls toward the
    center, as seen at t = 60 kyr.  Then the radiation pressure
    continues to force the gas outwards, increasing gas velocites up
    to 50 km/s.}
\end{figure}

Radiation pressure affects gas dynamics in an \ion{H}{2} region when
its force is comparable to the acceleration created by gas pressure of
the heated region.  The imparted acceleration on a hydrogen atom
$\mathbf{a}_{\rm rp} = E_{\rm ph}/c$.  This is especially important
when the ionization front is in its initial R-type phase.  Thus we
construct a test that focuses on a small scales, compared to the
Str\"{o}mgren radius.  The domain has a size of 8 pc with a uniform
density $\rho = 2900 \cubecm$ and initial temperature $T = 10^3$ K.
The source is located at the origin with a luminosity $L = 10^{50}$ ph
s$^{-1}$ and a $T=10^5$ K blackbody spectrum.  We use one energy group
$E_{\rm ph} = 29.6$ eV.  The grid is adaptively refined on overdensity
with the same criteron as Test 8.  The simulation is run for 150 kyr.

>>>>

Figure \ref{fig:test7_1} shows cuts of density, temperature, neutral
fraction, and pressure in a line connecting the source and the clump
center at $t = 1$, 10, and 50 Myr.  At 1 Myr, the ionization front has
propagated through the left-most 500 pc of the clump.  This heated gas
is now overpressurized, as seen in the pressure plot in Figure
\ref{fig:test7_1}, and then expands into the ambient medium.  This
expansion creates a photo-evaporative flow, seen in many star forming
regions \citep[e.g. M16][]{Hester96} as stars irradiate
nearby cold, dense overdensities.  These flows become evident in the
density at 10 Myr, seen both in the line cuts and slices (Figure
\ref{fig:test7_2}).  These outflows have temperatures up to 50,000 K.
At this time, the front has progressed about halfway through the
clump, if one inspects the neutral fraction.  However the high energy
photons have heated all but the rear surface of the clump.  At the end
of the test ($t = 50$ Myr), only the core and its associated shadow is
neutral, as seen in Figure \ref{fig:test7_3}.  The core has been
compressed by the surrounding warm medium, thus causing the higher
densities seen at $t = 50$ Myr.  The non-spherical artifacts on the
inner boundary of the warm outermost shell are caused by the initial
discretization of the sphere, as discussed in \S\ref{sec:test3}.

@@

Some astrophysical systems produce collimated radiation either
intrinisically by relavistic beaming or by an optically-thick torus
absorbing radiation in the equatorial plane.  The latter case would be
applicable in a subgrid model of active galactic nuclei (AGN) or
protostars, for example.  Simulating collimated radiation with ray
tracing is trivially accomplished by only initializing rays that are
within some opening angle $\theta_c$.

@@

We illustrate the expansion of the \ion{H}{2} region created by the
beamed radiation in Figure \ref{fig:test11}.  Before $t = 3$ Myr, the
\ion{H}{2} region is conical and contained within the dense clump,
depicted in the $t = 0.1$ Myr snapshot of the system.  At this time,
the ionization front is transitioning from R-type to D-type in the
transverse direction of the cone.  This can be seen in the minute
overdensities on the \ion{H}{2} transverse surface.  When it breaks
out of the overdensity, it creates a champange flow, where the
ionization front transitions back to a weak R-type front.  The cloud
surface is a constant-pressure contact discontinuity (CD) with a
density jump of 100.  After the front heats the gas at the CD, there
exists a pressure difference of $\sim 100$.  In response, the high
density gas accelerates into the ambient medium and heats it to $3
\times 10^4$ K.  Additionally a rarefaction wave travels towards the
clump center.  At later times, the transverse D-type front continues
through the clump, eventually creating a disk-like structure at the
final time.  The polar champange flows proceed to flow outwards,
creating a density shell with a diffuse ($10^{-28}$ \cubecm) and warm
(5000 K) medium in its wake.

@@

\begin{figure}[t]
  \epsscale{1.15}
  \plottwo{fig/test12_kph}{fig/test12_radial}
  \caption{\label{fig:test12_kph} Test 12 (Time variations of the
    source luminosity).  Left: Slice of the photo-ionization rate
    \kph~through the origin.  The source has a duty cycle of 0.5 Myr,
    and the box has a light crossing time of 3.3 Myr.  The shells of
    high \kph~originate from radiation that was emitted when the
    source was at its peak luminosity, illustrating the
    time-dependence of the radiative transfer equation.  Right: Radial
    profile \kph~with the inverse square law overplotted.}
  \epsscale{1}
\end{figure}

Our implementation retains the time derivative of the radiative
transfer equation (Eq. \ref{eqn:rteqn}) if we choose a constant ray
tracing timestep, which saves the photon packages between timesteps if
the $c\;dt_{\rm P} < L_{\rm box}$.  This effect only becomes apparent
when the variation timescale of the point source is smaller than the
light crossing time of the simulation.  Furthermore, the timestep
should resolve the variation timescale by a few times.  This property
might be important in large box simulations with variable sources,
e.g. AGN radiative feedback.  To test this, we can use an
exponentially varying source with some duty cycle.  In a functional
form, this can be described as
%
\begin{equation}
  \label{eqn:test13}
  L(t) = L_{\rm max} \times \exp[A(t_f/t_0 - 1)]
\end{equation}
%
where $t_f = 2 \times |t - t_0 \times \mathrm{round}(t/t_0)|$, $t_0$
is the duty cycle, and $A=4$ controls the width of the radiation
pulse.  To illustrate the effects of source varability, we remove any
dependence on the medium by considering an optically-thin uniform
density $\rho 10^{-4}~\cubecm$.  We take $L_{\rm box} = 1$ Mpc, which
has a light crossing time of 3.3 Myr.  A source is placed at the
origin with $L_{\rm max} = 10^{55}$ ph s$^{-1}$ and $t_0 = 0.5$ Myr.
We use a radiative transfer timestep of 50 kyr to resolve the duty
cycle by 10 timesteps.  The simulation is run for 6 Myr, so the
radiation propagates throughout the box.

The variability of the source is clearly illustrated in the
photo-ionization rates, shown in Figure \ref{fig:test12_kph}.  The
shells of relative maximum \kph~corresponds to radiation that was
emitted when the source was at its peak luminosity.  They are
separated by $ct_0$ Mpc and are geometrically diluted with increasing
radius.  Averaged over shells of the same width, photo-ionization
rates decrease as $1/r^2$.

Next we test the hydrodynamical response to a varying source by
repeating Test 5.  We set the peak luminosity $L_{\rm max} = 2 \times
10^{49}$ erg s$^{-1}$ that is a factor of 4 more luminous than Test
5, so the average luminosity is $\sim 5 \times 10^{48}$ erg s$^{-1}$.

@@

As the \ion{H}{2} region grows the magnetized medium, shown in Figures
\ref{fig:test14_1} and \ref{fig:test14_2}, it transforms from
spherical to oblate as it is magnetically confined in directions
perperdicular to the magnetic field.  This occurs at $t > 0.5$ Myr
because the magnetic pressure exceeds the thermal pressure, and the
gas can only flow along field lines.  \citeauthor{Krumholz07_ART}
observed some carbuncle artifacts along the ionization front; whereas
we see smooth density gradients, which is most likely caused by our
more diffusive choice of the HLL Riemann solver when compared to Roe's
Riemann solver used in \citet{Krumholz07_ART}, who also use PLM as a
reconstruction method.  The evolution of the magnetic field lines
evolve in a similar manner as their results.

@@

\begin{figure}[t]
  \epsscale{0.6}
  \plotone{fig/xray_profiles.eps}
  \caption{\label{fig:xray_fig} Radial profiles of temperature and
    ionized fraction showing the effects of secondary ionizations from
    a mono-chromatic 1 keV spectrum.  The discontinuity at $r \sim
    0.4$ is caused by artifacts in the ray tracing, which is described
    in Test 1.  The high energy photons can ionize multiple hydrogen
    atoms, increasing the ionized fraction.  In part, less radiation
    goes into thermal energy, lowering the temperature.}
  \epsscale{1}
\end{figure}

@@

Figure \ref{fig:xray_fig} shows the main effects of secondary
ionizations from the 1 keV spectrum on the ionization and thermal
state of the system.  Without secondary ionizations, each absorption
results in one ionization with the remaining energy transferred into
thermal energy.  But with secondary ionizations, recall that most of
the radiation energy goes into hydrogen and helium ionizations in
neutral gas; whereas in ionized gas, most of the energy is
thermalized.  In this test, only the inner 300 pc is completely
ionized because of the small cross-section of hydrogen at $E_{\rm ph}
= 1$ keV.  Beyond this core, the medium is only partially ionized.
This process expands the hot $T = 10^5$ K core by a factor of 2.  In
the outer neutral regions, the ionization fraction is larger by a
factor of $\sim 10$, which in turn results in less photo-heating,
lowering the temperature by a factor 2--3.

@@

\begin{figure}[t]
  \epsscale{1.075}
  \plottwo{fig/noRP_profiles.eps}{fig/RP_profiles.eps}
  \caption{\label{fig:rp_profiles} (a) No radiation pressure.
    Radial profiles of (clockwise from top left) density, temperature,
    radial velocity, and neutral fraction.  Time units in the legned
    are in kyr.  (b) Radial profiles with radiation pressure.  The
    momentum transferred to the gas drives out the gas at higher
    velocities than without radiation pressure.  Afterwards the
    central region is underpressurized, and the gas infalls toward the
    center, as seen at t = 60 kyr.  Then the radiation pressure
    continues to force the gas outwards, increasing gas velocites up
    to 50 km/s.}
  \epsscale{1}
\end{figure}

Radiation pressure affects gas dynamics in an \ion{H}{2} region when
its force is comparable to the acceleration created by gas pressure of
the heated region.  The imparted acceleration on a hydrogen atom
$\mathbf{a}_{\rm rp} = E_{\rm ph}/c$.  This is especially important
when the ionization front is in its initial R-type phase, where the
gas has not reacted to the thermal pressure yet.  Thus we construct a
test that focuses on a small scales, compared to the Str\"{o}mgren
radius.  The domain has a size of 8 pc with a uniform density $\rho =
2900 \cubecm$ and initial temperature $T = 10^3$ K.  The source is
located at the origin with a luminosity $L = 10^{50}$ ph s$^{-1}$ and
a $T=10^5$ K blackbody spectrum.  We use one energy group $E_{\rm ph}
= 29.6$ eV.  The grid is adaptively refined on overdensity with the
same criteron as Test 8.  The simulation is run for 140 kyr.

We compare nearly identical simulations but one with radiation
pressure and one without radiation pressure to quantify its effects.
Radial profiles of density, temperature, neutral fraction, and radial
velocity are shown in Figure \ref{fig:rp_profiles} for both
simulations.  Without radiation pressure, the evolution of the
\ion{H}{2} is matches the analytical expectations described in
\S\ref{sec:test5}.  At $t=140$ kyr, the ionization front radius is
increased by $\sim5\% = 0.16$ pc.  However radiation pressure impacts
the system the greatest inside the ionization front.  At $t = 40$ kyr,
the central density is smaller by a factor of 20 with radiation
pressure, but the temperatures are almost equal.  A rarefaction wave
thus propagates toward the center, depicted by the negative radial
velocities at $t = 60$ kyr.  This raises the central density to
$10^{-21}$ g \cubecm~at $t = 80$ kyr.  Afterwards, the radiation
continues to force gas outwards.  From $t = 100$ kyr to $t = 140$ kyr,
the maximum radial velocity of the ionized gas increases from 10
\kms~to 50 \kms.  This leaves behind an even more diffuse medium,
lowering the gas density by a factor of 10 at $t = 140$ kyr.  Thus the
recombination rates are lower, resulting in increased ionization
fractions and temperatures in the \ion{H}{2} region.
